Challenging Fuel Cycle Modeling Assumptions: Facility and Time Step Discretization Effects

### 2015-10-13 paper.tex
Add comment on how to measure fuel sharing effect
add equations for shortage and fuel sharing

    \item Groups of facilities (usually all facilities of a given type) are
        treated as a single, aggregate entity. Therefore, all facilities experience
        identical operational histories.

    \item Resource flows are modeled as continuous, within the limitations of time
        step discretization.  Therefore, material is acquired and discharge in
        uniform amounts at every time step.

@@

This work investigates in detail some differences between fleet vs. individual
reactor modeling.  The difference between these two modeling styles in terms
of their effects on output of fuel cycle simulators in general is not well
known. This analysis is designed to provide both quantitative and
qualitative insight about how fleet-based continuous flow models differ from
individual facility discrete flow models.

@@

    \item \emph{Case MI}: \underline{M}onthly time steps with \underline{I}ndividual reactor modeling
    \item \emph{Case MF}: \underline{M}onthly time steps with \underline{F}leet reactor modeling
    \item \emph{Case QI}: \underline{Q}uarterly (3-month) time steps with \underline{I}ndividual reactor modeling
    \item \emph{Case QF}: \underline{Q}uarterly (3-month) time steps with \underline{F}leet reactor modeling

@@

@@

A fleet reactor archetype was created for Cyclus.  Although the modeling of
individual, discrete facilities was a motivation for Cyclus, its flexibility
makes it relatively straightforward to create plug-in archetypes to model
anywhere along the spectrum from individual, discrete facilities to
continuous flow fleet facilities.  In order to contrast well with the individual,
single-facility granularity modeling that is part of the standard Cyclus archetypes, 
a fleet archetype was created for this work with aggregate facilitiy behavior and
continuous flow.

@@

    \item Reactor capacity is deployed and retired in increments of single
      reactor units, each with capable of producing power $P_r$ (e.g. 1 GWe
      LWRs).  $N_{r}$ reactors are deployed at any time capable of producing
      $P_r \cdot N_r$ power.

    \item In the event of fuel shortages, the number of reactors that can
      operate is modeled as

      \begin{equation}
        N_{o} = N_{r} \cdot \frac{C_{inv}}{C_{cap}},
        \label{eqn:operating-reactors}
      \end{equation}

       where $C_{inv}$ is the total amount of fuel in all reactor cores (kg),
       and $C_{cap}$ is the total fuel capacity for all reactor cores (kg).

@@

        \begin{equation}
            D =
            \frac{B}{L} \cdot N_{o} =  \frac{B}{L} \cdot N_{r}\cdot \frac{C_{inv}}{C_{cap}}
            \label{eqn:fleet-discharge}
        \end{equation}

        $D$ is the discharge rate (kg per time step), $B$ is the batch size
        (kg) for a single reactor's batch, $L$ is the cycle length (including
        refueling time) in time steps.  In the event of fuel shortages, the
        discharge rate is reduced because some reactors are not operating.

@@

    \item Generated power at any timestep, $P_G$, is based on the number of
      operating reactors:

      \begin{equation}
        P_{G} = P_r \cdot N_o = P_r \cdot N_r \cdot \frac{C_{inv}}{C_{cap}}
      \end{equation}

@@

The chosen scenario approximates a transition from a the current once-through light water
reactor (LWR) fleet to a full sodium fast reactor (SFR) fuel cycle within evaluation group 23 (EG-23)
in the DOE's Advanced Nuclear Fuel Cycle Options report
\cite{wigeland_nuclear_2014}. Scenario and transition details were patterned
after ongoing work by the DOE in their Fuel Cycle Options campaign.
%% Not sure if we should make such an explicit reference to the ongoing FCO
%% transition work until they publish something. On one hand, it reveals
%% things they are secretive about. On the other hand, we should give them due
%% credit.

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.47\textwidth]{exp2/flow.eps}
    \caption[Material flow paths]{Potential material flow paths between facility types.}
    %% consider a Cyclist-derived visualization here?
    \label{fig:flow}
\end{figure}

%% is it worth waiting until the 1.4 release?  Citing an ambiguous pre-release makes
%% this a little harder to reproduce.
A pre-release of Cyclus version 1.4 was used for this analysis, with the
standard archetypes provided with Cyclus in the included Cycamore
library - also a pre-release of version 1.4.  A custom storage archetype was
developed because Cycamore did not yet provide one.  The standard Cycamore reactor archetype
was used for both LWR and SFR reactor types in cases MI and QI.  Correspondingly
the custom fleet-based reactor archetype described earlier was used for both reactor
types in cases MF and QF.  Additional configuration for the reactor types is
shown in Table \ref{tab:reactors} (cycle length includes the refueling outage).

@@

\begin{table}[h]
    \caption{Reactor Facility Parameters}
    \centering
    \begin{tabular}{ |r | c c | }
        \hline                       
            & LWR & SFR \\
        \hline                       
        Lifetime (yr)         & 80 & 80 \\
        Cycle length (months) & 18 & 14 \\  %% why list this SFR cycle length if it's never used?
        Batch size (kg)       & 29565 & 7490 \\
        Batches per core      & 3 & 5 \\
        \hline                       
    \end{tabular}
    \captionsetup{justification=centering}
    \label{tab:reactors}
\end{table}

The described scenario is used as closely as possible in each of the four
cases.  However, a few small case-dependent changes to the scenarios relating to
reactor configuration are necessary.  For case \emph{QI}, 
cycle and outage times must be adjusted to be a
multiple of 3 months. 
%% this tweak is a distraction.  Why not simply declare the SFR cycle length &
%% outage as 12/3 from the beginning.  Is this because the fuel recipes you
%% are using assume the other parameters and you don't want to sweep that
%% under the rug?
 In order to make the comparison more valid, the cases
with monthly time steps are also adjusted to have identical cycle and outage
times.  The invariants preserved with respect to reactor behavior between all
four cases are shown in Table \ref{tab:invar}.  Table \ref{tab:reactor-detail} shows
the computed/selected configuration for both LWR and SFR reactor types for all
four cases.

@@

\begin{table*}
    \centering
    \captionsetup{justification=centering}
    \caption{Reactor Parameters by Case}
    \begin{tabular}{ |r | c c | c c | }
        \hline                       
                                          & \multicolumn{2}{c |}{LWR}       & \multicolumn{2}{c |}{SFR} \\
                                          & Cases MI,QI & Cases MF,QF (fleet) & Cases MI,QI & Cases MF,QF (fleet)  \\
        \hline                       
        Cycle length (months)             & 15        & 18                & 12        & 15 \\
        Refueling outage (months)         & 3         & 0                 & 3         & 0 \\
        Batch size ($\text{kg} \cdot \text{HM}$) & 29,565    & 88,695            & 8,025     & 40,125 \\
        Power capacity (MWe)              & 1,080     & 900               & 450       & 360 \\
        \hline                       
    \end{tabular}
    \label{tab:reactor-detail}
\end{table*}

In each of the four cases, the initial 100 LWRs are modified to have a zero
length refueling outage with an explicit 900 MWe net power output capacity.  Because the
individual reactor model doesn't have any way to start a reactor mid-cycle,
this was done to avoid having all initial LWRs refuel at the same time.  The
parameters for each case were also selected to keep capacity factors for
individual reactors very roughly around 80\%.

@@

One difference between the fleet and individual facility modeling is the
\emph{quantized shutdown effect}.  If a fleet-based reactor modeling 3-batch
cores is short one half batch of fuel, it will produce $\frac{2.5}{3}\cdot P_r$
of power for that time step where $P_r$ is the power capacity of a single
reactor of the fleet (\emph{e.g.}~1000~MWe).  If an individually modeled reactor is
short a half batch of fuel, then it will produce no power ($0 \cdot P_r$) that
time step.  Time step duration can amplify this effect.  For both fleet and
individual reactor modeling, power goes offline in full time step increments.
Consider an example of a 3-batch reactor with a fuel shortage of $\frac{1}{2}$ batch of
fuel that would be resolve in 1 month.  Depending on the modeling choice, each reactor
would produce the following amounts of energy over a 3 month period:

\begin{itemize}
  \item case \emph{QF}: $\frac{2.5}{3} \cdot 3 \text{mo} \cdot P_r = 2.5P_r$
  \item case \emph{MF}: $[\frac{2.5}{3} \cdot 1 \text{mo} + 1 \cdot 2
    \text{mo}] \cdot P_r = 2.83P_r$
  \item case \emph{QI}: $0 \cdot 3 \text{mo} \cdot P_r = 0$
  \item case \emph{MI}: $[0\cdot 1 \text{mo} + 1 \cdot 2\text{mo}] \cdot P_r = 2 P_r$
\end{itemize}

Quantifying this effect directly is difficult.  One way of observing
it is to compare fleet reactor scenarios that have different time step
durations (i.e. cases MF and QF).  Because the fleet reactors have perfect fuel
sharing (described below) and no noise from refueling outages, differences observed will
likely be due to quantized shutdown.

@@

During the initial 100 years, cases MI and QI have larger variance in power
output than cases MF and QF.  This is caused by minor refueling cycle
synchronization.  The \emph{cycle staggering effect} becomes much more visible
during and after the fuel shortage as evidenced by the larger swings in power
generation from time step to time step.  Initially, the reactor cycles are
staggered well.  During the shortage, several reactors that previously had
staggered cycles are all offline together waiting for fuel.  On time steps
where reactors retire (not just during the shortage), new deployments are made
to replace them in addition to new deployments made to address power demand
growth.  Because the initial reactors retire in waves somewhat close together,
there are corresponding waves of deployment, and these waves are echoed every
80 years (the reactor lifetime).  These waves of deployment cause surges in
recycled fuel availability when they begin to discharge their fuel
%% these need to be LWR waves, right, or they'd just make things worse?
that cause many of the reactors offline during the
shortage to receive fuel and come online together.  The net effect is that the
fuel shortage degrades reactor cycle stagering overall.  As the simulation
continues, however, new reactors continue to be deployed and reactors with
synchornized cycles retire resulting in a gradual return to better
staggering visible by the end of the 200 years.

>>>>

    \item Groups of facilities (usually all facilities of a given type) are
        treated as a single, aggregate entity.

    \item Resource flows are often (approximately) continuous (limited by time
        step discretization).

@@

This work investigates in detail some differences between fleet vs. individual
reactor modeling.  The difference between these two modeling styles in terms
of their affects on output of fuel cycle simulators in general is not well
known. This analysis is designed to provide both quantitative and
qualitative insight about how fleet-based continuous flow models differ from
individual facility discrete flow models.

@@

    \item \emph{Case MI}: Monthly time steps with individual reactor modeling
    \item \emph{Case MF}: Monthly time steps with fleet reactor modeling
    \item \emph{Case QI}: Quarterly (3-month) time steps with individual reactor modeling
    \item \emph{Case QF}: Quarterly (3-month) time steps with fleet reactor modeling

@@

The case labels consist of two letters each: "M" for monthly time steps, "Q"
for quarterly time steps, "I" for individual reactors, and "F" for fleet
reactors.

@@

A fleet reactor archetype was created for Cyclus.  The flexibility of Cyclus
makes it relatively straight forward to create plug-in archetypes to model
anywhere along the spectrum from individual, discrete facilities to
continuous flow fleet facilities.  In order to contrast well with individual,
single-facility granularity modeling, the fleet archetype created here falls more
toward the aggregate, continuous flow side of the spectrum.

@@

    \item Reactor capacity is deployed and retired in single reactor units
        (e.g. 1 GWe LWRs).

@@

        \begin{equation}
            D =
            \frac{B}{L} \cdot N_{r} \cdot \frac{C_{inv}}{C_{cap}}
            \label{eqn:fleet-discharge}
        \end{equation}

        $D$ is the discharge rate (kg per time step), $B$ is the batch size
        (kg) for a single reactor's batch, $L$ is the cycle length (including
        refueling time) in time steps, $N_{r}$ is the number of reactors in
        the fleet, $C_{inv}$ is the total amount of fuel in all reactor cores
        (kg), and $C_{cap}$ is the total fuel capacity for all reactor cores
        (kg).  The last fractional term accounts for lower fuel burning rates
        due to unfueled reactors.

@@

    \item Generated power is computed similarly to core discharge rates by
        multiplying the total fleet power capacity by the same last fractional
        term of Equation \ref{eqn:fleet-discharge} to account for unfueled
        capacity in the fleet.

@@

The scenario modeled approximates a transition from a the current once-through light water
reactor (LWR) fleet to a full sodium fast reactor (SFR) fuel cycle within evaluation group 23 (EG-23)
in the DOE's Advanced Nuclear Fuel Cycle Options report
\cite{wigeland_nuclear_2014}. Scenario and transition details were patterned
after ongoing work by the DOE in their Fuel Cycle Options campaign.

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.47\textwidth]{exp2/flow.eps}
    \caption[Material flow paths]{Potential material flow paths between facility types.}
    \label{fig:flow}
\end{figure}

A pre-release of Cyclus version 1.4 was used for this analysis.  Archetypes
used were the standard ones provided with Cyclus in the included Cycamore
library - also a pre-release of version 1.4.  A custom storage archetype was
used because Cycamore did not yet provide one.  The standard cycamore reactor archetype
was used for both LWR and SFR reactor types in cases MI and QI.  Correspondingly
the custom fleet-based reactor archetype described earlier was used for both reactor
types in cases MF and QF.  Additional configuration for the reactor types is
shown in Table \ref{tab:reactors} (cycle length includes the refueling outage).

@@

\begin{table}[h]
    \caption{Reactor Facility Parameters}
    \centering
    \begin{tabular}{ |r | c c | }
        \hline                       
            & LWR & SFR \\
        \hline                       
        Lifetime (yr)         & 80 & 80 \\
        Cycle length (months) & 18 & 14 \\
        Batch size (kg)       & 29565 & 7490 \\
        Batches per core      & 3 & 5 \\
        \hline                       
    \end{tabular}
    \captionsetup{justification=centering}
    \label{tab:reactors}
\end{table}

The described scenario is used as closely as possible in each of the four
cases.  However, a few small case-dependent changes to the scenarios relating to
reactor configuration are necessary.  For the cases using quarterly time steps
with individual reactors, cycle and outage times must be adjusted to be a
multiple of 3 months.  In order to make the comparison more valid, the cases
with monthly time steps are also adjusted to have identical cycle and outage
times.  The invariants preserved with respect to reactor behavior between all
four cases are shown in Table \ref{tab:invar}.  Table \ref{tab:reactor-detail} shows
the computed/selected configuration for both LWR and SFR reactor types for all
four cases.

@@

\begin{table*}
    \centering
    \captionsetup{justification=centering}
    \caption{Reactor Parameters by Case}
    \begin{tabular}{ |r | c c | c c | }
        \hline                       
                                          & \multicolumn{2}{c |}{LWR}       & \multicolumn{2}{c |}{SFR} \\
                                          & Cases MI,QI & Cases MF,QF (fleet) & Cases MI,QI & Cases MF,QF (fleet)  \\
        \hline                       
        Cycle length (months)             & 15        & 18                & 12        & 15 \\
        Refueling outage (months)         & 3         & 0                 & 3         & 0 \\
        Batch size ($\text{kg} \cdot \text{HM}$) & 29,565    & 29,565            & 8,025     & 40,125 \\
        Power capacity (MWe)              & 1,080     & 900               & 450       & 360 \\
        \hline                       
    \end{tabular}
    \label{tab:reactor-detail}
\end{table*}

In each of the four cases, the initial 100 LWRs are modified to have a zero
length refueling outage with an explicit 900 MWe net power output capacity.  Because the
individual reactor model doesn't have any way to start a reactor mid-cycle,
this was done to avoid having all initial LWRs refuel at the same time.  The
parameters for each case were also selected to keep capacity factors for
individual reactors very roughly around 90\%.

@@

One difference between the fleet and individual facility modeling is the
\emph{quantized shutdown effect}.  If a fleet-based reactor modeling 3-batch
cores is short one half batch of fuel, it will produce $\frac{2.5}{3}\cdot C$
of power for that time step where $C$ is the power capacity of a single
reactor of the fleet (e.g. 1000 MWe).  If an individually modeled reactor is
short a half batch of fuel, then it will produce no power ($0 \cdot C$) that
time step.  Time step duration can also have this effect.  For both fleet and
individual reactor modeling, power goes offline in full time step increments.
For example, if a 3-batch fleet based reactor is short $\frac{1}{2}$ batch of
fuel with a 3 month time step, then it will produce $\frac{2.5}{3} \cdot 3
\text{mo} \cdot C $ of energy.  In contrast, a monthly time step might allow
the same reactor to produce $[\frac{5}{6} \cdot 1 \text{mo} + 1 \cdot 2
\text{mo}] \cdot C $ of energy if the fuel shortage actually ended after only
1 month.  Quantifying this effect directly is difficult.  One way of observing
it is to compare fleet reactor scenarios that have different time step
durations (i.e. cases MF and QF).  Because the fleet reactors have perfect fuel
sharing (described below) and no noise from refueling outages, differences observed will
likely be due to quantized shutdown.

@@

During the initial 100 years, cases MI and QI have larger variance in power
output than cases MF and QF.  This is caused by minor refueling cycle
synchronization.  The \emph{cycle staggering effect} becomes much more visible
during and after the fuel shortage as evidenced by the larger swings in power
generation from time step to time step.  Initially, the reactor cycles are
staggered well.  During the shortage, several reactors that previously had
staggered cycles are all offline together waiting for fuel.  On time steps
where reactors retire (not just during the shortage), new deployments are made
to replace them in addition to new deployments made to address power demand
growth.  Because the initial reactors retire in waves somewhat close together,
there are corresponding waves of deployment, and these waves are echoed every
80 years (the reactor lifetime).  These waves of deployment cause waves in
recycled fuel availability that cause many of the reactors offline during the
shortage to receive fuel and come online together.  The net effect is that the
fuel shortage degrades reactor cycle stagering overall.  As the simulation
continues, however, new reactors continue to be deployed and reactors with
synchornized cycles retire resulting in a gradual return to better
staggering visible by the end of the 200 years.
